---
title: 'Analyst Intern, Data Science & Solutions Project'
output: html_document
author: "Chikunie, Praise"
date: "`r format(Sys.Date(), '%m/%d/%y')`"
---

```{r set options, include=FALSE}
# DO NOT CHANGE THE LINE BELOW 
knitr::opts_chunk$set(echo = TRUE)
```

``` {css styling, echo=FALSE}

<style>
.tocify {
max-width: 175px !important;
}
</style>

<style>
.main-container {
width: 100%;
max-width: 940px;
margin-left: 250px;
margin-right: auto;
}
</style>

<style>
.red-header {
  color: red;
}
</style>

```

```{r logo, echo = FALSE}

htmltools::img(src = 'https://cdn.nba.com/logos/nba/1610612760/primary/L/logo.svg',
                height = '250px',
                alt = 'logo',
                style = 'position: fixed; top: -40px; left: 5px;')
```


# Introduction  

The purpose of this project is to gauge your technical skills and problem solving ability by working through something similar to a real NBA data science project. You will work your way through this R Markdown document, answering questions as you go along. Please begin by adding your name to the "author" key in the YAML header. When you're finished with the document, come back and type your answers into the answer key at the top. Please leave all your work below and have your answers where indicated below as well. Please note that we will be reviewing your code so make it clear, concise and avoid long printouts. Feel free to add in as many new code chunks as you'd like.

Remember that we will be grading the quality of your code and visuals alongside the correctness of your answers. Please try to use the tidyverse as much as possible (instead of base R and explicit loops.)  

**Note:**    

**Throughout this document, any `season` column represents the year each season started. For example, the 2015-16 season will be in the dataset as 2015. For most of the rest of the project, we will refer to a season by just this number (e.g. 2015) instead of the full text (e.g. 2015-16).**   

<h1 class="red-header">Answers</h1>  

## Part 1      

**Question 1:**   

- 1st Team: XX.X points per game  
- 2nd Team: XX.X points per game  
- 3rd Team: XX.X points per game  
- All-Star: XX.X points per game   

**Question 2:** XX.X Years  

**Question 3:** 

- Elite: X players.  
- All-Star: X players.  
- Starter: X players.  
- Rotation: X players.  
- Roster: X players.  
- Out of League: X players.  

**Open Ended Modeling Question:** Please show your work and leave all responses below in the document.


## Part 2  

**Question 1:** XX.X%   
**Question 2:** Written question, put answer below in the document.    
**Question 3:** Written question, put answer below in the document.    
  

# Setup and Data    

```{r load data, message = F, warning = F}
library(tidyverse)
# Note, you will likely have to change these paths. If your data is in the same folder as this project, 
# the paths will likely be fixed for you by deleting ../../Data/awards_project/ from each string.
awards <- read_csv("awards_data.csv")
player_data <- read_csv("player_stats.csv")
team_data <- read_csv("team_stats.csv")
rebounding_data <- read_csv("team_rebounding_data_22.csv")

```

## Part 1 -- Awards  

In this section, you're going to work with data relating to player awards and statistics. You'll start with some data manipulation questions and work towards building a model to predict broad levels of career success.  

### Question 1  

**QUESTION:** What is the average number of points per game for players in the 2007-2021 seasons who won All NBA First, Second, and Third teams (**not** the All Defensive Teams), as well as for players who were in the All-Star Game (**not** the rookie all-star game)?

```{r}
nba_n_star <- awards %>%
  filter(season >= 2007 & season <= 2021) %>%
  filter(
    all_nba_points_rk %in% c(5, 6, 7) |
      all_star_game == "TRUE"
  )%>%
  select(
    "season", "nbapersonid", "all_nba_points_rk", "all_star_game", "All NBA First Team", 
    "All NBA Second Team", "All NBA Third Team")

```

```{r}
# Selecting only the necessary columns from player_data

game_points <- player_data %>%
  select(nbapersonid, points)

# Merge the filtered awards data with the selected player data
point_data <- nba_n_star %>%
  left_join(game_points, by = "nbapersonid")

str(point_data)
```

```{r}

# Calculate the average points per game for each group
average_1st_team <- mean(point_data$points[point_data$`All NBA First Team` == "1"], na.rm = TRUE)
average_2nd_team <- mean(point_data$points[point_data$`All NBA Second Team` == "1"], na.rm = TRUE)
average_3rd_team <- mean(point_data$points[point_data$`All NBA Third Team` == "1"], na.rm = TRUE)
average_all_star <- mean(point_data$points[point_data$all_star_game == "TRUE"], na.rm = TRUE)
```

```{r}
# the results
cat("1st Team:", round(average_1st_team, 1), "points per game\n")
cat("2nd Team:", round(average_2nd_team, 1), "points per game\n")
cat("3rd Team:", round(average_3rd_team, 1), "points per game\n")
cat("All-Star:", round(average_all_star, 1), "points per game\n")
View(point_data)


# Calculating the average points per game
average_points_ <- mean(point_data$points, na.rm = TRUE)

cat("Average Points Per Game:", average_points_)
```



<span style="color:red">**ANSWER 1:**</span>  

1st Team: XX.X points per game  
2nd Team: XX.X points per game  
3rd Team: XX.X points per game  
All-Star: XX.X points per game   


### Question 2  

**QUESTION:** What was the average number of years of experience in the league it takes for players to make their first All NBA Selection (1st, 2nd, or 3rd team)? Please limit your sample to players drafted in 2007 or later who did eventually go on to win at least one All NBA selection. For example:

- Luka Doncic is in the dataset as 2 years. He was drafted in 2018 and won his first All NBA award in 2019 (which was his second season).  
- LeBron James is not in this dataset, as he was drafted prior to 2007.  
- Lu Dort is not in this dataset, as he has not received any All NBA honors.  


```{r}
# Filtering for players drafted in 2007 or later who won All NBA selection
nba_select <- awards %>%
  filter(
    season >= 2007,
    (all_nba_points_rk %in% c(5, 6, 7))
  )

# Grouping by nbapersonid and calculating the minimum season they won NBA selection
first_nba_season <- nba_select %>%
  group_by(nbapersonid) %>%
  summarize(first_nba_season = min(season))
```

```{r}
# Selecting only necessary columns from player data
player_draft <- player_data %>%
  distinct(nbapersonid, .keep_all = TRUE) %>%
  select(nbapersonid, draftyear, points, player)

# Merge with the player data
data_merged <- first_nba_season %>%
  left_join(player_draft, by = "nbapersonid")
View(data_merged)
```

```{r}
# Calculating the years of experience to make the first All NBA selection
data_merged$years_to_nba <- data_merged$first_nba_season - data_merged$draftyear

```

```{r}
# Printing the result for each player
for (i in 1:nrow(data_merged)) {
  player_name <- data_merged[i, "player"]
  years_to_nba <- data_merged[i, "years_to_nba"]
  draft_year <- data_merged[i, "draftyear"]
  first_nba_year <- data_merged[i, "first_nba_season"]
  
  if (years_to_nba == 1) {
    message <- paste(player_name, "is in the dataset as", years_to_nba, "year. He was drafted in", draft_year,
                     "and won his first All NBA award in", first_nba_year, ".")
  } else if (years_to_nba > 1) {
    message <- paste(player_name, "is in the dataset as", years_to_nba, "years. He was drafted in", draft_year,
                     "and won his first All NBA award in", first_nba_year, ".")
  } else {
    message <- paste(player_name, "is not in this dataset, as he has not received any All NBA honors.")
  }
  
  cat(message, "\n")
}
```


<span style="color:red">**ANSWER 2:**</span>  

XX.X Years  


## Data Cleaning Interlude  

You're going to work to create a dataset with a "career outcome" for each player, representing the highest level of success that the player achieved for **at least two** seasons *after his first four seasons in the league* (examples to follow below!). To do this, you'll start with single season level outcomes. On a single season level, the outcomes are:  

- Elite: A player is "Elite" in a season if he won any All NBA award (1st, 2nd, or 3rd team), MVP, or DPOY in that season.    
- All-Star: A player is "All-Star" in a season if he was selected to be an All-Star that season.   
- Starter:  A player is a "Starter" in a season if he started in at least 41 games in the season OR if he played at least 2000 minutes in the season.    
- Rotation:  A player is a "Rotation" player in a season if he played at least 1000 minutes in the season.   
- Roster:  A player is a "Roster" player in a season if he played at least 1 minute for an NBA team but did not meet any of the above criteria.     
- Out of the League: A player is "Out of the League" if he is not in the NBA in that season.   

We need to make an adjustment for determining Starter/Rotation qualifications for a few seasons that didn't have 82 games per team. Assume that there were 66 possible games in the 2011 lockout season and 72 possible games in each of the 2019 and 2020 seasons that were shortened due to covid. Specifically, if a player played 900 minutes in 2011, he **would** meet the rotation criteria because his final minutes would be considered to be 900 * (82/66) = 1118. Please use this math for both minutes and games started, so a player who started 38 games in 2019 or 2020 would be considered to have started 38 * (82/72) = 43 games, and thus would qualify for starting 41. Any answers should be calculated assuming you round the multiplied values to the nearest whole number.

Note that on a season level, a player's outcome is the highest level of success he qualifies for in that season. Thus, since Shai Gilgeous-Alexander was both All-NBA 1st team and an All-Star last year, he would be considered to be "Elite" for the 2022 season, but would still qualify for a career outcome of All-Star if in the rest of his career he made one more All-Star game but no more All-NBA teams. Note this is a hypothetical, and Shai has not yet played enough to have a career outcome.   

Examples:  

- A player who enters the league as a rookie and has season outcomes of Roster (1), Rotation (2), Rotation (3), Roster (4), Roster (5), Out of the League (6+) would be considered "Out of the League," because after his first four seasons, he only has a single Roster year, which does not qualify him for any success outcome.  
- A player who enters the league as a rookie and has season outcomes of Roster (1), Rotation (2), Starter (3), Starter (4), Starter (5), Starter (6), All-Star (7), Elite (8), Starter (9) would be considered "All-Star," because he had at least two seasons after his first four at all-star level of production or higher.  
- A player who enters the league as a rookie and has season outcomes of Roster (1), Rotation (2), Starter (3), Starter (4), Starter (5), Starter (6), Rotation (7), Rotation (8), Roster (9) would be considered a "Starter" because he has two seasons after his first four at a starter level of production. 

### Question 3  

**QUESTION:** There are 73 players in the `player_data` dataset who have 2010 listed as their draft year. How many of those players have a **career** outcome in each of the 6 buckets?  

```{r}
#loading the career dataset
career_data <- read_csv("career_dataset.csv")

# Filtering players with draftyear = 2010
players_draft_year_2010 <- player_data %>%
  filter(draftyear == 2010) %>%
  distinct(nbapersonid)

# Join with career dataset  
year_2010_outcomes <- players_draft_year_2010 %>%
  left_join(career_data, by = "nbapersonid") %>%
  group_by(career_outcome) %>%
  summarise(count = n())

# Filling NAs with 0 
year_2010_outcomes$count[is.na(year_2010_outcomes$count)] <- 0


career_outcomes <- unique(year_2010_outcomes$career_outcome)

for (outcome in career_outcomes) {
  if(!is.na(outcome)) {
    outcome_count <- sum(year_2010_outcomes$career_outcome == outcome, na.rm = TRUE)
    cat(outcome, ": ", outcome_count, " players.\n")
  }
}

```

<span style="color:red">**ANSWER 3:**</span>    

Elite: X players.  
All-Star: X players.  
Starter: X players.  
Rotation: X players.  
Roster: X players.  
Out of League: X players.  

### Open Ended Modeling Question   

In this question, you will work to build a model to predict a player's career outcome based on information up through the first four years of his career. 

This question is intentionally left fairly open ended, but here are some notes and specifications.  

1. We know modeling questions can take a long time, and that qualified candidates will have different levels of experience with "formal" modeling. Don't be discouraged. It's not our intention to make you spend excessive time here. If you get your model to a good spot but think you could do better by spending a lot more time, you can just write a bit about your ideas for future improvement and leave it there. Further, we're more interested in your thought process and critical thinking than we are in specific modeling techniques. Using smart features is more important than using fancy mathematical machinery, and a successful candidate could use a simple regression approach. 

2. You may use any data provided in this project, but please do not bring in any external sources of data. Note that while most of the data provided goes back to 2007, All NBA and All Rookie team voting is only included back to 2011.  

3. A player needs to complete at least three additional seasons after their first four to be considered as having a distinct career outcome for our dataset. (We are using 3+ instead of 2+ just to give each player a little more time to accumulate high level seasons before we classify his career). Because the dataset in this project ends in 2021, this means that a player would need to have had the chance to play in the '21, '20, and '19 seasons after his first four years, and thus his first four years would have been '18, '17, '16, and '15. **For this reason, limit your training data to players who were drafted in or before the 2015 season.** Karl-Anthony Towns was the #1 pick in that season.  

4. Once you build your model, predict on all players who were drafted in 2018-2021 (They have between 1 and 4 seasons of data available and have not yet started accumulating seasons that inform their career outcome).  

5. You can predict a single career outcome for each player, but it's better if you can predict the probability that each player falls into each outcome bucket.    

6. Include, as part of your answer:  
  - A brief written overview of how your model works, targeted towards a decision maker in the front office without a strong statistical background. 
  - What you view as the strengths and weaknesses of your model.  
  - How you'd address the weaknesses if you had more time and or more data.  
  - A ggplot or ggplotly visualization highlighting some part of your modeling process, the model itself, or your results.  
  - Your predictions for Shai Gilgeous-Alexander, Zion Williamson, James Wiseman, and Josh Giddey.  
  - (Bonus!) An html table (for example, see the package `reactable`) containing all predictions for the players drafted in 2019-2021.  


```{r}
library(dplyr)
library(nnet)
library(ggplot2)
library(tidyr)

# Load the datasets
awards <- read.csv("awards_data.csv")
player_data <- read.csv("player_stats.csv")
career_dataset <- read.csv("career_dataset.csv")


# Selecting relevant columns from player_stats
selected_player_data <- player_data %>%
  select(
    season, nbapersonid, team,
    games, games_start, mins, fgm, fga,
    fgp, fgm3, fga3, fgp3, fgm2, fga2,
    fgp2, efg, ftm, fta, ftp,
    off_reb, def_reb, tot_reb,
    ast, steals, blocks, tov, tot_fouls,
    points, PER, FTr, off_reb_pct,
    def_reb_pct, tot_reb_pct, ast_pct,
    stl_pct, blk_pct, tov_pct, usg,
    OWS, DWS, WS, OBPM, DBPM, BPM, VORP
  )

# Remove duplicates from player_stats
player_stats <- selected_player_data %>%
  distinct(season, nbapersonid, .keep_all = TRUE)

# Merge the cleaned datasets
merged_data <- awards %>%
  left_join(player_stats, by = c("season", "nbapersonid"))

career_data <- career_dataset %>%
  distinct(nbapersonid, .keep_all = TRUE)

merged_data <- merged_data %>%
  left_join(career_data, by = "nbapersonid")

# Define your outcome variable (e.g., All-Star selection)
outcome_variable <- "career_outcome"

# Define the features you want to use for modeling
features <- c(
  "points", "ast", "tot_reb",
  "blocks", "stl_pct", "off_reb_pct",
  "def_reb_pct", "ast_pct", "blk_pct","fgm", 
  "mins", "fgm3", "fga3", "fgp3"
)

# Prepare the data for modeling
model_data <- merged_data %>%
  select(all_of(c(outcome_variable, features))) %>%
  na.omit()


# Example assuming your outcome variable is in a column named "career_outcome"
model_data$career_outcome <- as.integer(factor(model_data$career_outcome))

set.seed(123)
train_indices <- sample(nrow(model_data), 0.8 * nrow(model_data))
train_data <- model_data[train_indices, ]
test_data <- model_data[-train_indices, ]

model <- multinom(career_outcome ~ ., data = train_data)

predictions <- predict(model, newdata = test_data)
# Example: Confusion Matrix
conf_matrix <- table(test_data$career_outcome, predictions)


# Create a data frame with predicted probabilities
probabilities <- predict(model, newdata = test_data, type = "probs")
probabilities_df <- as.data.frame(probabilities)
colnames(probabilities_df) <- levels(factor(test_data$career_outcome))

# Add the actual outcome to the data frame
probabilities_df$Actual <- factor(test_data$career_outcome)

# Create a mapping of numeric class labels to meaningful class names
class_names <- c("Elite", "All-Star", "Starter", "Rotation", "Roster", "Out of League")

# Replace numeric class labels with meaningful class names
probabilities_df$Actual <- factor(probabilities_df$Actual, levels = 1:6, labels = class_names)

# Select the "Elite" class for visualization
class_to_visualize <- "Elite"

# Reshape the data for plotting using gather
probabilities_long <- gather(probabilities_df, key = "Class", value = "Probability", -Actual)

# Create a bar plot for the "Elite" class and other classes
ggplot(probabilities_long, aes(x = Class, y = Probability, fill = Actual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Predicted Class", y = "Probability") +
  scale_fill_manual(name = "Actual Class", values = setNames(c("blue", "green", "red", "purple", "orange", "gray"), class_names)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle(paste("Predicted Probabilities for", class_to_visualize, "Class"))


```

## Part 2 -- Predicting Team Stats  

In this section, we're going to introduce a simple way to predict team offensive rebound percent in the next game and then discuss ways to improve those predictions.  
 
### Question 1   

Using the `rebounding_data` dataset, we'll predict a team's next game's offensive rebounding percent to be their average offensive rebounding percent in all prior games. On a single game level, offensive rebounding percent is the number of offensive rebounds divided by their number offensive rebound "chances" (essentially the team's missed shots). On a multi-game sample, it should be the total number of offensive rebounds divided by the total number of offensive rebound chances.    

Please calculate what OKC's predicted offensive rebound percent is for game 81 in the data. That is, use games 1-80 to predict game 81.  

```{r}
# Load data
rebounding_data <- read.csv("team_rebounding_data_22.csv")
View(rebounding_data)
# Filter to just OKC rows 
okc <- rebounding_data[rebounding_data$team == "OKC",]

# Take rows for games 1-80
okc_first80 <- okc[1:80,] 

# Sum offensive rebounds and chances
okc_orb <- sum(okc_first80$offensive_rebounds)
okc_orch <- sum(okc_first80$off_rebound_chances)

# Calculate offensive rebound percentage
okc_orb_pct <- okc_orb / okc_orch

# Print predicted percentage 
print(okc_orb_pct)
```

<span style="color:red">**ANSWER 1:**</span>    

XX.X% 

### Question 2  

There are a few limitations to the method we used above. For example, if a team has a great offensive rebounder who has played in most games this season but will be out due to an injury for the next game, we might reasonably predict a lower team offensive rebound percent for the next game.  

Please discuss how you would think about changing our original model to better account for missing players. You do not have to write any code or implement any changes, and you can assume you have access to any reasonable data that isn't provided in this project. Try to be clear and concise with your answer.  

<span style="color:red">**ANSWER 2:Here are a few ways I think would improve the model to account for missing players:

- Include player-specific offensive rebounding percentages. Consider the team forecast based on which players will miss the upcoming game and each player's season offensive rebound rate.


- Determine the minimal impact of missing players using lineup data. Examine the differences in offensive rebounding rates between teams when a major player is on and off the court. Adjust the projection to account for any missing lineup combinations.


- Develop a regression model based on team and player statistics. As predictors, use a regression with team rebounding metrics as well as individual player rebounding rates, minutes, and so on. When forecasting, leave out the absent player's variables.


- Make use of some adjusted plus-minus. Estimate each player's contribution to team rebounding when they are on the court. Reduce the team estimate by the plus-minus of the absent players.


- Using player monitoring data, train a machine learning model. Learn spatial patterns and the impacts of player absence by training on optical tracking data. Make projections for the next game without missing player data.


Rather than assuming the team aggregate as fixed, the idea is to identify the impact of missing players by position and use and then update the team estimate accordingly. More player or lineup data would enable advanced modifications.

**</span>  


### Question 3  

In question 2, you saw and discussed how to deal with one weakness of the model. For this question, please write about 1-3 other potential weaknesses of the simple average model you made in question 1 and discuss how you would deal with each of them. You may either explain a weakness and discuss how you'd fix that weakness, then move onto the next issue, or you can start by explaining multiple weaknesses with the original approach and discuss one overall modeling methodology you'd use that gets around most or all of them. Again, you do not need to write any code or implement any changes, and you can assume you have access to any reasonable data that isn't provided in this project. Try to be clear and concise with your answer.  

<span style="color:red">**ANSWER 3:Here are a few other potential flaws in using the average offensive rebound % to forecast future performance:

1. Regression to the mean - Teams may experience "lucky" or "unlucky" games that cause them to diverge from their genuine talent level. Taking the raw average alone will outweigh these anomalies. A Bayesian method that regresses to the population mean may enhance projections.

2. Schedule impacts - The strength of the schedule varies during the season. A team's averages may momentarily fall due to harder opponents. This could be addressed by incorporating opponent adjustments.

3. Roster changes - Trades, injuries, and other personnel changes can all have an impact on rebounding abilities. A model that takes into consideration roster composition and minutes played could capture personnel effects.

A Bayesian hierarchical team rating model would suffice to address these difficulties. It would estimate an underlying genuine rebounding talent by using the team's rebounding percentage from each game, adjusted for opponent strength and roster changes. It would project new games based on estimated talent, adjusted for opponent, roster, and regression to the mean. To generate more accurate estimates, this methodology adjusts for outlier games, schedule effects, and changing personnel.
**</span>    




